---
title: 'NOAA Storm Database: Health and Economic Impacts'
author: "Kevin Joy"
date: "10/15/2019"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Executive Summary

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.  

The data analysis addresses the following questions:

1. Across the United States, which types of events (as indicated in the \color{red}{\verb|EVTYPE|}EVTYPE variable) are most harmful with respect to population health?

2. Across the United States, which types of events have the greatest economic consequences?


## Data Analysis:

### 1. Download Data, clean and prepocess data
- Load necessary packages
```{r}
library(tidyverse)

```

- Download data csv file does not exist unzip activity.zip
```{r cache=TRUE}
data_url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
dataName <- "data/storm.csv.bz2"
if(!dir.exists("data")){dir.create("data")}


if(!file.exists(dataName)){
    download.file(data_url, dataName)
    #no need to unzip
  }
storm = read.csv(dataName, nrows = 10000)
str(storm)
```


### 2. Exploratory Data Analysis

## Findings


